Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2025-01-17 03-methods.qmd
revpaper -  theory, methods & conclusions
methods & descriptives

The main data source is the Chilean Longitudinal Social Survey 2016–2023. The ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile. The sampling design is probabilistic, stratified, clustered, and multistage. It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities. Consequently, the ELSOC comprises a total of 2,927 participants aged between 18 and 75 years in wave 1. In addition, the sample represents 77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The waves 2016, 2017, 2018, 2019 and 2022 were administered using computer-assisted personal interviewing (CAPI). However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021. In addition, wave 3 included a refreshment sample to counter survey attrition. The same sampling strategy as that of wave 1 was implemented for selecting new cases. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in the analytical sample because we wanted to analyze longer response trends. Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants. In broader terms, the accumulated attrition between wave 1 and wave 7 is 40%, achieving a final sample of N = 1741. The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/. The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.

@@

For testing the hypothesis regarding status we included educational level and household income. Also, we included subjective social status as it has been argued that perceived social status is a relevant predictor of attitudes toward economic inequality [@schneider_poverty_2015;@castillo_meritocracia_2019]. In addition, subjective measures can complement objective measures in predicting life chances [@oesch_subjective_2023], which are arguably connected to how individuals experience economic inequality. In this regard, it has been argued that how people form their views on their standing in society results from experiences with direct socioeconomic circumstances and social comparison processes with reference groups [@Condon2020]. Finally, political identification on a left-right scale, age groups, and gender were included as control variables. @tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.

>>>>

The main data source is the Chilean Longitudinal Social Survey ELSOC (2016–2023). ELSOC has been designed as a yearly panel study to evaluate how individuals think, feel, and behave regarding a set of social issues concerning conflict and social cohesion in Chile. The sampling design is probabilistic, stratified, clustered, and multistage. It provides adequate coverage of the country's largest cities (Metropolitan Area of Santiago, Valparaíso, and Concepción) and smaller cities. The sample covers participants aged between 18 and 75 years (in wave 1), representing  77% of the total national population and 93% of the urban areas [@elsoc_estudio_2022].

The survey has been conducted yearly since 2016, except for the year 2020, when it was suspended due to the pandemic. The waves 2016, 2017, 2018, 2019, 2022, and 2023 were administered using computer-assisted personal interviewing (CAPI). However, a reduced version was conducted using computer-assisted telephone interviews (CATI) in 2021. In addition, wave 3 included a refreshment sample to counter survey attrition. As a result, the total sample of wave 3 included 3,748 cases, of which 2,229 were part of the original sample and 1,519 were from the refreshment sample. The data from the refreshment sample is not included in the analytical sample for this study as we analyze longer response trends. Regarding the original sample, the response rate was 62.4% in wave 1, achieving N = 2,927 participants. In broader terms, the accumulated attrition between wave 1 and wave 7 is 40%, achieving a final sample of N = 1,741. The analysis used longitudinal weights to avoid bias due to the sampling design, allowing us to control for biases arising from systematic patterns of non-participation in the survey after the first wave. For a more detailed analysis of responses, attrition, and the construction of longitudinal weights, visit https://coes.cl/encuesta-panel/. The dataset is publicly available at: https://dataverse.harvard.edu/dataverse/elsoc.

@@

For testing the hypothesis regarding status we considered educational level and household income. Also, we included subjective social status as it has been argued that perceived social status is a relevant predictor of attitudes toward economic inequality [@schneider_poverty_2015;@castillo_meritocracia_2019]. In addition, such subjective measures can complement objective measures in predicting life chances [@oesch_subjective_2023], which are arguably connected to how individuals experience economic inequality. In this regard, it has been argued that how people form their views on their standing in society results from experiences with direct socioeconomic circumstances and social comparison processes with reference groups [@Condon2020]. Finally, political identification on a left-right scale, age groups, and gender were included as control variables. @tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.

### 2025-01-17 04-analisis.qmd
methods & descriptives

@@

@fig-alluvial illustrates yearly frequencies in the justification of educational inequality between 2016 and 2023. Each year represents stacked percentual frequencies, and the flows in between reflect the within-subject change of opinions from one year to the next, as we are using longitudinal panel data [@rosvall_mapping_2010]. For instance, of the 32.9% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories. In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last waves. This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11% in 2023).

@@

## Multivariate analysis

@tbl-multilevel shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.167, p<.001), 2018 ($\beta$=-0.059, p<.05), and 2019 ($\beta$=-0.03, p>.05). Nevertheless, in the last waves of 2022 and 2023 there is a radical increase in level of justification of educational inequality ($\beta$=0.204, p<.001 and $\beta$=0.233, p<.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time, but on the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.

>>>>

Beginning with a descriptive analysis of changes in the justification of educational inequality between 2016 and 2023, @fig-alluvial illustrates yearly frequencies and transitions between years. Each year represents stacked porcentual frequencies, whereas the flows in between reflect the within-subject change of opinions from one year to the next [@rosvall_mapping_2010]. For instance, of the 32.9% who strongly disagreed with inequality justification in 2016, about half of them kept responding the same in 2017, whereas the other half shifted their opinion to other response categories. In general, the large majority - between 80 and 90% - disagrees with the justification of educational inequality throughout the years. Despite this overall tendency, we also observe that the disagreement with inequality justification (disagree + strongly disagree) tends to go down in the last waves. This change is mostly a result of the increase in the "agree" category, which doubles compared to previous years (from 5.4% in 2016 to 11% in 2023).

@@

@@

_Multilevel models_

@tbl-multilevel shows the multilevel estimation results for the justification of educational inequality. Model 1 includes the survey waves to estimate intertemporal changes in the dependent variable. Taking 2016 as a reference point, we can observe a staggered decrease in 2017 ($\beta$=-0.167, p<.001), 2018 ($\beta$=-0.059, p<.05), and 2019 ($\beta$=-0.03, p>.05). Nevertheless, in the last waves of 2022 and 2023 there is a clear increase in level of justification of educational inequality ($\beta$=0.204, p<.001 and $\beta$=0.233, p<.001), suggesting a non-linear change in this variable. Attempting to model this path of change over time, Model 2 incorporates time (survey waves) as a continuous variable as well as its quadratic term, representing the nonlinear association initially observed in Model 1. On the one hand, the survey wave depicts a negative association, expressing an average decrease in inequality justification over time. On the other hand, the quadratic wave term is positive, indicating the reversion of this path in the last measurements point.
